\NAME: A framework for \\ Efficient Intrusion Detection\\ in the Internet of Things

### 2014-12-08 paper.tex
fixed typo’s, sentences, semi-colons,…

When intrusion prevention fails, intrusion detection serves as a second layer
of defense. It looks for patterns and anomalies that can indicate malicious
behavior. However, supporting an adequate set of intrusion detection algorithms
imposes significant overhead on resource-constrained devices such as those in
the Internet of Things. Manual fusion of the algorithms reduces resource
consumption by optimizing resource sharing, yet, proves to be time-consuming,
repetitive and error-prone. To address this problem we propose \NAME, a
framework for the development of efficient intrusion detection systems. It
consists of a domain specific language and a code generator. The language
allows formally describing the intent of intrusion detection algorithms and
supports the generator in organizing the source code. A side-by-side comparison
shows that \NAME-generated code reduces message passing overhead, execution
time and memory footprint in comparison to sequential calls to individual
implementations of the algorithms.

@@

In wired networks, firewalls focus on the outer perimeter of the network,
filtering unwanted packets and protecting the entire internal network. But
attacks on flaws in services can pass unnoticed. This is where intrusion
detection (ID) comes into play: an intrusion detection system (IDS) monitors
all traffic that passes through the firewall, looking for patterns of malicious
activity, and optionally, after detecting such pattern, alerts the firewall,
allowing it to take corrective actions \cite{denning1987intrusion}.

@@

The internet of things (IoT) holds great potential to positively influence our
daily work and life through domotics, assisted living, e-health and enhanced
learning \cite{atzori2010internet}. But with this potential the IoT also
presents a significant threat: by opening our smart homes and personal data to
all of these interconnected devices, we also open them to everyone who is able
to break the virtual locks that protect them (i.e.\ passwords \& encryption
keys). Implementation of security measures by itself is hindered by the fact
that IoT devices typically have limited batteries, processing power and memory.

@@

Current developments in ID on resource-constrained devices focus on programming
frameworks that structure the implementation of algorithms \cite{valero2012di}
and offer the required basic functional components to implement them
\cite{krontiris2008lidea}. A key shortcoming of prior ID frameworks, is that
they don't offer a way to optimize the usage of resources. A solution that
supports the integration of multiple algorithms on IoT devices needs to avoid
accumulating their impact on the available resources. A second omission is
support for systematic reuse to create different configurations. Addressing
this in a transparent and automated way offers the potential to significantly
reduce the effort of developing an IDS.

@@

The first contribution of this paper consists of the formulation of a design
pattern for ID algorithms, based on properties identified from a classification.

@@

The remainder of this paper proceeds as follows: section \ref{problems}
describes the inherent problems with the combination of ID algorithms. Section
\ref{classification} analyses and classifies ID algorithms. Section
\ref{pattern} looks for patterns in the implementation of these classes.
Section \ref{design} describes the design we applied to construct \NAME, its
domain specific language (DSL) and code generator. Section \ref{evaluation}
discusses and evaluates an implementation. Section \ref{related} explores
related work in the field of DSLs and code generation. Finally, section
\ref{conclusion} summarizes our findings, draws conclusions and identifies
topics for future work.

@@

The combination of multiple ID algorithms tends to result in suboptimal code.
Besides these code-level problems, operational problems also hinder the
implementation of an IDS for IoT devices.

@@

  \item[Loops] by themselves are not a problem, but different loops, that
  perform the same operations in sequence, can incur overhead. An example is
  the handling of messages: each algorithm parses each observed message on the
  network, processing the same byte stream, performing comparable operations to
  find similar patterns.

@@

  \item[The network] allows algorithms to operate on a distributed scale. But
  scattered calls to the a message sending function cause multiple accesses to
  the wireless radio, keeping it active and energy consuming more than strictly
  needed.

@@

Research into ID in wireless networks typically focuses on two major topics:
the devices as single entities and the network as a group of such devices. Both
are important: a group cannot make a decision without members that detect
malicious behavior and a group-based decision is often needed because devices
in the network can miss certain events that could reveal intrusions. In a
wireless network, where not all participants are within each others' radio
range, wireless devices typically only have a partial view on the entire
network. This prohibits them from observing potentially crucial events in
distant parts of the network. The fact that resource-constrained devices
typically apply duty cycling to lower their energy consumption, further
increases this problem, introducing periods of time where they cannot observe
local events.

@@

There are different ways to construct a taxonomy for intrusion detection, but
common themes do appear. According to \cite{mishra2004intrusion} and
\cite{ioannis2007towards} there are three major categories: 1) anomaly
detection, 2) signature or misuse detection, and 3) specification-based
detection. In \cite{alrajeh2013intrusion} the authors agree with this typology
and add the notion of hybrid intrusion detection systems and cross layer
intrusion detection systems as recent advances in research. Although these
additional categories seem promising, the authors admit that the impact on
resource-constrained devices will be too much. We therefore focus on the three
categories agreed upon by most sources.

@@

  \item[Signature or misuse detection] starts with known patterns of attacks.
  It tries to find evidence for intrusions in the actions of and communication
  with a device. Because attacks often use valid ways to communicate with
  services on the devices, this approach can produce false positives.

@@

Not all algorithms can decide if an attack is real or not, based on a single
event. Signature and misuse detection can in some cases, but anomaly detection
typically deals with some level of uncertainty. Also, in distributed setups it
is impossible to fully trust any neighboring device. In these situations one
has to rely on statistics and thresholds to decide when to drop trust in a
given device.

@@

The previous sections identified loops, common data and the network as three
important code-level problems. These problems are inherent to ID algorithms.
But at the same time they conflict with the target environment, the IoT\@.
\NAME provides a framework to efficiently implement an IDS for the IoT and
offers a solution for the identified problems. It consists of a 1) DSL, 2)
source code generator and 3) software library. Figure \ref{fig:design}
visualizes the overall design.

@@

The introduction of a code generator in the development process changes the
traditional way of working. Most importantly because the actual source code is
no longer the driving force that can be controlled by developers. It is
literally an intermediate representation between code generator and compiler.

A code generator constructs this representation using three components, as
shown in figure \ref{fig:design}: 1) the description of the algorithm in the
DSL, 2) information about the domain, in this case the Nodes domain and 3)
information about the platform and target language. At the level of the
compiler, a software library is introduced to offload specialized
implementations and allow custom extensions.

@@

This hierarchy is important from a development effort point of view. The level
of dependency on the targeted platform dictates the level of effort to
implement changes to that platform. For example: to generate a different
language will require a complete rewrite of the software library and a partial
rewrite of the language description of the generator. Although this is a
reusable, one-time effort, it is important in relation to the scalability of
the framework itself.

@@

This section therefore discusses \NAME's approach to 1) loops, 2) nodes domain,
3) unified messaging, 4) language support and illustrates these with a complete
example.

@@

Most, if not all, ID algorithms deal with information about other devices in
the network. To share this information \NAME provides a unified \ttt{nodes}
concept as part of its DSL and exposes it as an object to the implementer.

Algorithms can extend nodes with their own properties. This way, only one
instance for every known device will be present in memory. Finally, nodes can
define a scope for a function to execute in. This way, nodes offer a way to
i.e.\ iterate all known devices, again without the need for an explicit loop.

@@

Access to the network, scattered around an algorithm, causes the wireless radio
to be more active than needed. Delaying the sending of messages and grouping
them, can lower this overhead.

@@

The semantic model is the core of the solution. It offers a way to describe ID
algorithms in a way that allows semantic reinterpretation through functional
code fusion. This way, reorganizing the functionality of different algorithms
allows for more optimal execution and memory usage. Hence the name \NAME of the
framework: intrusion detection functionality organization optimization.

@@

A software library allows offloading of common implementation aspects, thus
avoiding small discrepancies in code and suboptimal use of shared resources.
Section \ref{dsl-unified-msg} illustrated this approach with a case for unified
messaging: parsing of incoming messages and grouping of outgoing messages.

Other common functionality isn't generated as is either. The execution of
functions based on a schedule or the implementation of higher level conceptual
functions, such as \ttt{now} and \ttt{sha1}, are calls into the standard
software library.

@@

\NAME reduces inefficiencies in network access, execution and memory usage,
that occur due to a selection of ID algorithms. To achieve this, it avoids
constructions that are hard to identify and reorganize, such as loops and
common data. A DSL, focussing on event-handling, enforces these restrictions
and enables formally describing ID algorithms.

@@

In this section we present a prototype implementation and evaluate its
performance. We first present the implementation consisting of the hardware
platform, network and software stack. We first take an analytical approach and
look at the handling of common data and the size of the source code. Next, we
use the implementation to quantify its merits using a side-by-side comparison
of generated code versus a manual implementation. We consider experimental
results regarding usage of the network, execution time of the event-loop and
the image size.

@@

Relying on basic hardware and software allows us to validate that the generator
is capable of generating code even for the most basic environment. It shows
that the requirements of the generator towards the platform are minimal, don't
rely on advanced frameworks or operating systems and therefore are applicable
in any environment, targeting any platform.

@@

We started from a basic application that measures light intensity. The
application was an includable piece of code, which allowed integration in a
manual implementation and simple merging when generation. On top of this
baseline we added two intrusion detection related algorithms: a
watchdog \cite{mishra2004intrusion}, that allows nodes to validate each other's
presence, and a reputation-based algorithm \cite{ganeriwal2008reputation}, that
checks if a parent node is cooperative and actually forwards messages further
upstream.

@@

Looking at the generated source code is a straightforward way to validate the
framework. A simple metric is comparing the amount of code versus the size of
the formal description of ID algorithms. This can be an indication of the
development effort and complexity. In this experiment, implementing both
algorithms manually results in 637 lines of C code. The corresponding \NAME
implementation results in 156 lines of DSL code. This is largely thanks to the
higher level coding constructs, as introduced in section \ref{language-support}
dealing with \NAME's language support.

Next we look at data structures. Sharing data eliminates redundant information,
such as a network address, from the scope of each individual algorithm.
Analysis of the source code with both algorithms implemented, shows that this
reduces the memory footprint for one node entry by 2 bytes, from 27 bytes to 25
bytes. In our implementation we used the ZigBee network address, which indeed
consists of 16 bits. This results in a reduction of 7.4\%. Using ZigBee's 64
bits unique addresses, this reduction can easily become 20.5\%.

@@

Besides this analysis we also instrumented the code and evaluated four
criteria: the network usage in number of packets and bytes, the time required
to perform once cycle of the event loop and the image size of the resulting
compiled code. To do this, we collected these metrics for four situations:
without ID algorithms, with a watchdog, with reputation-tracking and with both
algorithms implemented.

@@

Table \ref{tbl:results} shows the collected data for the manual and the
generated implementation, as well as a comparison in absolute values and
relative.

These results show that adding algorithms manually to the base-application has
a cumulative effect. The impact of both algorithms is simply the sum of the
impact of each algorithm by itself. The execution time of one cycle of the
event loop with both algorithms enabled is even higher than simply the sum.

@@

The first observation deals with the base case: both the manual and the
generated implementations perform exactly the same. There is no fundamental
structural difference in the overall architecture of the code.

Secondly, when turning to the goal to reduce the network usage, we see that
both the number of packets as well as the amount of bytes sent have dropped by
13\% and 7\%.

Thirdly, focusing on the execution time, both algorithms by themselves add an
equal amount of time to the processing of a single event loop cycle. But when
combined, the event loop is about 7\% faster compared to the manual case.

Finally, we see that the software library that comes with the generated code
adds to the size of the resulting image. But the cost is close to constant and
is lower for both algorithms together. With roughly 3KB of additional overhead,
or 15\% in this simple case, the impact is small. Taking into account the
available memory on the ATMega1284p, this is only 2.3\% of the available
resources.

@@

TinyScript \cite{levis2004tinyscript} is an example of a language for wireless
sensor networks. It's a complete programing language that aims to ease the
description of algorithms in general. The Basic-like syntax does enable more
productive development. But this approach doesn't address the underlying
resource problems inherently connected to the context.

SEAL \cite{elsts2013seal} is a novel complete programming language that focuses
on novice programmers. It is a clear fit for \emph{sense-and-send} type of
applications. It does however share the idea of events to overcome the technical
nature of loops.

@@

Optimizing for power efficiency however proves to be concurrent to size and
performance optimization, and requires different strategies. Data and memory
optimizations \cite{panda2001data} can result in improvements on a lower level.
But really preserving energy requires higher-level strategies, targeting
architecture, radio handling, efficient software,\dots \cite{naik2001software}.
Most of these are beyond the control of the compiler and require a higher level
of abstraction to address them.

@@

Applying code generation to ID, often focusses on the generation of native code
to avoid the interpretation of descriptions of patterns to look for. I2S
\cite{charitakis2003code} is an example that generates code for SNORT
\cite{roesch1999snort} rules. To optimize this highly efficient code, it often
targets a specific platform.

@@

Adequate intrusion detection for IoT devices requires multiple algorithms to
operate at the same time. Frequent reselection, e.g. due to changing
requirements or security threats, requires a flexible way to incorporate the
different algorithms.

@@

\NAME is a framework consisting of a DSL, a source code generator and software
library. Using the DSL, intrusion detection algorithms can be formally
described. The code generator applies functional code fusion to interpret the
intent of algorithms and merges common functionality and shared data. The
software library hosts common functionality, such as network message parsing
and event scheduling.

Initial experiments confirm the theoretical positive impact on resource usage:
grouping of messages lowers the number of packets by 13\%, while the total
number of bytes sent over the network is reduced by 7\%. The execution time is
improved by 7\% and memory usage can be reduced by even 20\%.

@@

As a next step, we will implement more ID algorithms using \NAME and continue
the experiments on a larger scale. This way we will also be able to verify the
initially observed inclination in the evolution of the image size with growing
numbers of algorithms.

Future research will focus on the integration of applications with a generated
IDS. We see opportunities for applications who detect intrusions, to actively
and dynamically respond and avoid malicious payloads.

>>>>

When intrusion prevention fails, intrusion detection serves as a second layer
of defense. It looks for patterns and anomalies that can indicate malicious
behavior. However, supporting an adequate set of intrusion detection algorithms
imposes significant overhead on resource-constrained devices such as those in
the Internet of Things. Manual fusion of the algorithms reduces resource
consumption by optimizing resource sharing, yet, proves to be time-consuming,
repetitive and error-prone. To address this problem we propose \NAME, a
framework for the development of efficient intrusion detection systems. It
consists of a domain specific language and a code generator. The language
allows formally describing the intent of intrusion detection algorithms and
supports the generator in organizing the source code. A side-by-side comparison
shows that \NAME-generated code reduces message passing overhead, execution
time and memory footprint in comparison to sequential calls into individual
implementations of the algorithms.

@@

In wired networks, firewalls focus on the outer perimeter of the network,
filtering unwanted packets and protecting the entire internal network. Still,
attacks on flaws in services can pass unnoticed. This is where intrusion
detection (ID) comes into play: an intrusion detection system (IDS) monitors
all traffic that passes through the firewall, looking for patterns of malicious
activity, and optionally, after detecting such pattern, alerts the firewall,
allowing it to take corrective actions \cite{denning1987intrusion}.

@@

The internet of things (IoT) holds great potential to positively influence our
daily work and life through domotics, assisted living, e-health and enhanced
learning \cite{atzori2010internet}. With this potential the IoT also presents a
significant threat: by opening our smart homes and personal data to all of
these interconnected devices, we also open them to everyone who is able to
break the virtual locks that protect them (i.e.\ passwords \& encryption keys).
Implementation of security measures by itself is hindered by the fact that IoT
devices typically have limited batteries, processing power and memory.

@@

Current developments in ID on resource-constrained devices focus on programming
frameworks that structure the implementation of algorithms \cite{valero2012di}
and offer the required basic functional components to implement them
\cite{krontiris2008lidea}. A key shortcoming of prior ID frameworks is that
they don't offer a way to optimize the usage of resources. A solution that
supports the integration of multiple algorithms on IoT devices needs to avoid
accumulating their impact on the available resources. A second omission is
support for systematic reuse to create different configurations. Addressing
this in a transparent and automated way offers the potential to significantly
reduce the effort of developing an IDS.

@@

The first contribution of this paper consists in the formulation of a design
pattern for ID algorithms, based on properties identified from a classification.

@@

The remainder of this paper proceeds as follows: section \ref{problems}
describes the inherent problems in the combination of ID algorithms. Section
\ref{classification} analyses and classifies ID algorithms. Section
\ref{pattern} looks for patterns in the implementation of these classes.
Section \ref{design} describes the design we applied to construct \NAME, its
domain specific language (DSL) and code generator. Section \ref{evaluation}
discusses and evaluates an implementation. Section \ref{related} explores
related work in the field of DSLs and code generation. Finally, section
\ref{conclusion} summarizes our findings, draws conclusions and identifies
topics for future work.

@@

The combination of ID algorithms tends to result in suboptimal code. Besides
these code-level problems, operational problems also hinder the implementation.

@@

  \item[Loops] by themselves are not a problem, but different loops, that
  perform the same operations in sequence, can cause overhead. An example is
  the handling of messages: each algorithm parses each observed message on the
  network, processing the same byte stream, performing comparable operations to
  find similar patterns.

@@

  \item[The network] allows algorithms to operate on a distributed scale.
  However, scattered calls to a message sending function cause multiple
  accesses to the wireless radio, keeping it active and energy consuming more
  than strictly needed.

@@

Research into ID in wireless networks typically focuses on two major topics:
the devices as single entities and the network as a group of such devices. Both
are important: a group cannot make a decision without members that detect
malicious behavior and a group-based decision is often needed because devices
in the network can miss certain events that could reveal intrusions. In a
wireless network, where not all participants are within each others' radio
range, wireless devices typically only have a partial view on the entire
network. This prohibits them from observing potentially crucial events in
distant parts of the network. The fact that resource-constrained devices
typically apply duty cycling to lower their energy consumption, further
increases this problem, introducing periods of time when they cannot observe
local events.

@@

There are different ways to construct a taxonomy for intrusion detection, but
common themes do appear. According to \cite{mishra2004intrusion} and
\cite{ioannis2007towards} there are three major categories: 1) anomaly
detection, 2) signature or misuse detection, and 3) specification-based
detection. In \cite{alrajeh2013intrusion} the authors agree with this typology
and add the notion of hybrid intrusion detection systems and cross layer
intrusion detection systems as recent advances in research. Although these
additional categories seem promising, the authors admit that the impact on
resource-constrained devices will be too high. We therefore focus on the three
categories agreed upon by most sources.

@@

  \item[Signature or misuse detection] starts with known patterns of attacks.
  It tries to find evidence of intrusions in the actions of and communication
  with a device. Because attacks often use valid ways to communicate with
  services on the devices, this approach can produce false positives.

@@

Not all algorithms can decide if an attack is real or not, based on a single
event. Signature and misuse detection can do that in some cases, but anomaly
detection typically deals with some level of uncertainty. Also, in distributed
setups it is impossible to fully trust any neighboring device. In these
situations one has to rely on statistics and thresholds to decide when to drop
trust in a given device.

@@

The previous sections identified loops, common data and the network as three
important code-level problems. These problems are inherent to ID algorithms,
but conflict with the target environment, the IoT\@. \NAME provides a framework
to efficiently implement an IDS for the IoT and offers a solution for the
identified problems. It consists of a 1) DSL, 2) source code generator and 3)
software library. Figure \ref{fig:design} visualizes the overall design.

@@

The introduction of a code generator in the development process changes the
traditional way of working. This is most importantly due to the fact that the
actual source code is no longer the driving force that can be controlled by
developers. It is literally an intermediate representation between code
generator and compiler.

A code generator constructs this representation using three components, as
shown in figure \ref{fig:design}: 1) the description of the algorithm in the
DSL, 2) information about the domain, in this case the Nodes domain, and 3)
information about the platform and target language. At the level of the
compiler, a software library is introduced to offload specialized
implementations and allow custom extensions.

@@

This hierarchy is important from a development effort point of view. The level
of dependency on the targeted platform dictates the level of effort to
implement changes in that platform. For example: to generate a different
language will require a complete rewrite of the software library and a partial
rewrite of the language description of the generator. Although this is a
reusable, one-time effort, it is important in relation to the scalability of
the framework itself.

@@

This section therefore discusses \NAME's approach to 1) loops, 2) nodes domain,
3) unified messaging, 4) language support, and illustrates these with a
complete example.

@@

Most, if not all, ID algorithms deal with information about other devices in
the network. To share this information, \NAME provides a unified \ttt{nodes}
concept as part of its DSL and exposes it as an object to the implementer.

Algorithms can extend nodes with their own properties. This allows for only one
instance per known device to be present in memory. Finally, nodes can define a
scope for a function to execute in. This way, nodes offer a way to i.e.\
iterate all known devices, again without the need for an explicit loop.

@@

Access to the network, scattered around an algorithm, causes the wireless radio
to be more active than needed. Delaying the sending of messages and grouping
them can lower this overhead.

@@

The semantic model is the core of the solution. It offers a way to describe ID
algorithms so that semantic reinterpretation through functional code fusion is
possible. This reorganization of the functionality of different algorithms
allows for more optimal execution and memory usage. Hence the name \NAME of the
framework: intrusion detection functionality organization optimization.

@@

A software library allows offloading of common implementation aspects, thus
avoiding small discrepancies in code and suboptimal use of shared resources.
Previously we illustrated this approach with a case for unified messaging:
parsing of incoming messages and grouping of outgoing messages.

More common functionality isn't generated verbatim either. The execution of
functions based on a schedule or the implementation of higher level conceptual
functions, such as \ttt{now} and \ttt{sha1}, are calls into the standard
software library.

@@

\NAME reduces inefficiencies in network access, execution and memory usage,
that occur due to a selection of ID algorithms. To achieve this, it avoids
constructions that are hard to identify and reorganize, such as loops and
common data. A DSL, focusing on event-handling, enforces these restrictions and
enables formally describing ID algorithms.

@@

In this section we present a prototype implementation and evaluate its
performance. We start by introducing the implementation, consisting of the
hardware platform, network and software stack. We then continue with an
analytical approach, looking at the handling of common data and the size of the
source code. Next, we use the implementation to quantify its merits using a
side-by-side comparison of generated code versus a manual implementation. We
consider experimental results regarding usage of the network, execution time of
the event-loop and image size.

@@

Relying on basic hardware and software allows us to validate that the generator
is capable of generating code even for the most basic environment. It shows
that the requirements of the generator towards the platform are minimal, don't
rely on advanced frameworks or operating systems, and therefore are applicable
in any environment, targeting any platform.

@@

We started from a basic application that measures light intensity. The
application was an includable piece of code, which allowed integration in a
manual implementation and easy merging when generating. On top of this baseline
we added two intrusion detection related algorithms: a watchdog
\cite{mishra2004intrusion}, that allows nodes to validate each other's
presence, and a reputation-based algorithm \cite{ganeriwal2008reputation}, that
checks if a parent node is cooperative and actually forwards messages further
upstream.

@@

Looking at the generated source code is a straightforward way to validate the
framework. A simple metric is comparing the amount of code versus the size of
the formal description of ID algorithms. This can be an indication of the
development effort and complexity. Implementing both algorithms manually
results in 637 lines of C code. The corresponding \NAME implementation results
in 156 lines of DSL code. This is largely thanks to the higher level coding
constructs, as introduced in section \ref{design}, dealing with \NAME's
language support.

Next we look at data structures. Sharing data eliminates redundant information,
such as a network address, from the scope of each individual algorithm.
Analysis of the source code with both algorithms implemented, shows that this
reduces the memory footprint for one node entry by 2 bytes, from 27 bytes to 25
bytes. In our implementation we used the ZigBee network address, which indeed
consists of 16 bits. This results in a reduction of 7.4\% of dynamic memory per
known device. Using ZigBee's 64 bits unique addresses, this reduction can
easily become 20.5\%.

@@

Besides this analytical approach, we also instrumented the code and evaluated
four criteria: the network usage in number of packets and bytes, the time
required to perform once cycle of the event loop and the image size of the
resulting compiled code. To do this, we collected these metrics for four
situations: without ID algorithms, with a watchdog, with reputation-tracking
and with both algorithms implemented.

@@

Table \ref{tbl:results} shows the collected data for the manual and the
generated implementation, as well as a comparison in absolute values and in
relative terms.

These results show that adding algorithms manually to the base-application has
a cumulative effect. The impact of both algorithms is simply the sum of the
impact of each algorithm by itself. The execution time of one cycle of the
event loop with both algorithms enabled is even higher than simply the sum of
their individual times.

@@

The first observation deals with the base case: both the manual and the
generated implementations perform in exactly the same way. There is no
fundamental structural difference in the overall architecture of the code.

Secondly, when turning to the goal of reducing the network usage, we see that
both the number of packets as well as the amount of bytes sent have dropped by
13\% and 7\%.

Thirdly, focusing on the execution time, both algorithms by themselves add an
equal amount of time to the processing of a single event loop cycle. When
combined, the event loop is about 7\% faster compared to the manual case.

Finally, we see that the software library that comes with the generated code
adds to the size of the resulting image. However, the cost is close to constant
and is lower for both algorithms together. With roughly 3KB of additional
overhead, or 15\% in this simple case, the impact is small. Taking into account
the available memory on the ATMega1284p, this is only 2.3\% of the available
resources.

@@

TinyScript \cite{levis2004tinyscript} is an example of a language for wireless
sensor networks. It's a complete programing language that aims to ease the
description of algorithms in general. The Basic-like syntax does enable more
productive development. Still, this approach doesn't address the underlying
resource problems inherently connected to the context.

SEAL \cite{elsts2013seal} is a novel complete programming language that focuses
on novice programmers. It is a clear fit for \emph{sense-and-send} type of
applications. It does however share the idea of using events to overcome the
technical nature of loops.

@@

Optimizing for power efficiency however proves to be concurrent to size and
performance optimization, and requires different strategies. Data and memory
optimizations \cite{panda2001data} can result in improvements on a lower level.
To really preserve energy one requires higher-level strategies, targeting
architecture, radio handling, efficient software,\dots \cite{naik2001software}.
Most of these are beyond the control of the compiler and require a higher level
of abstraction to address them.

@@

Applying code generation to ID often focusses on the generation of native code
to avoid the interpretation of descriptions of patterns to look for. I2S
\cite{charitakis2003code} is an example that generates code for SNORT
\cite{roesch1999snort} rules. To optimize this highly efficient code, it often
targets a specific platform.

@@

Adequate intrusion detection for IoT devices requires multiple algorithms to
operate at the same time. Frequent reselection, e.g. due to changing
requirements or security threats, requires a flexible way to incorporate the
algorithms.

@@

\NAME is a framework consisting of a DSL, a source code generator and software
library. Using the DSL, intrusion detection algorithms can be formally
described. The code generator applies functional code fusion to interpret the
intent of algorithms and merges common functionality \& shared data. The
software library hosts common functionality, such as network message parsing
and event scheduling.

Initial experiments confirm the theoretically positive impact on resource
usage: grouping of messages lowers the number of packets by 13\%, while the
total number of bytes sent over the network is reduced by 7\%. The execution
time is improved by 7\% and memory usage can be reduced by even 20\%.

@@

As a next step, we will implement more ID algorithms using \NAME and continue
experiments on a larger scale. This way we will also be able to verify the
initially observed inclination in the evolution of the image size with growing
numbers of algorithms.

Future research will focus on the integration of applications with a generated
IDS. We see opportunities for applications that detect intrusions, to actively
and dynamically respond and avoid malicious payloads.
